Fast Sorting on Modern Computers

### 2005-05-26 paper/docs/heap.tex
make way for the new and improved fs

>>>>

\section{Predictor results}

\sixplot{base_heapsort-0.eps}{cache4_heapsort-0.eps}{cache_heapsort-0.eps}{cache_heapsort-1.eps}{}{}{base and cache heapsorts}

The results of the software predictor show the relation of the branches
in the creation and destruction of the heap. The base heapsort results in
Figure \ref{base and cache heapsorts-1} show that the time taken to create the
heap is not significant relative to the time taken to sort from the heap.

The most interesting thing is the relation between the predictivity of the
comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
Figure \ref{Branch prediction performance for base and cache heapsorts}.
Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
8-heap, but has to make more comparisons. However, there are three comparisons
in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
comparisons in the 4-heap, and 150\% more in the 8-heap.

These branches become more predictable. The first predicts correctly 50\% of the
time, while the next predicts correctly 60\% of the time, then 71\%, 77\%, and
so on in a curve. This curve is shown in Figure TODO.
%TODO make a pretty graph for this.

Also notable from the results is that branches are more often not-taken then
they are correctly predicted, implying that a semi-static predictor based on
hints from the compiler would do a better job at making correct predictions.

### 2005-05-26 paper/docs/merge.tex
make way for the new and improved fs

\section{Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that $index(source) + index(auxiliary) =
C$.

@@

\section{Multi-mergesort}

Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the lists grow big, the overhead involved in this is
substantial. It is necessary to check each of the lists, and each key searched
would cause a cache miss, since each array segment is aligned, and map to
the same cache block.

@@

\section{Double Tiled Mergesort \\/ Double Multi-mergesort}

While writing the cache mergesorts, I realized several opportunities for
improvement. The first of these is avoiding the copy back to memory. In the
event of an odd number of merge steps, the auxiliary array will contain the
sorted array, which needs to be copied back. This can be avoided by pre-sorting
to make lists sized two or eight. A large increase of instruction count was realized by
sorting to lists of length two, so a length of eight was used. When an even number
was required, four was used.

The second improvement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since is the penalty
for level 1 misses is not great.

Finally, the merge is rewritten in a more straightforward and readable manner. This
was necessary due to the increased complexity of debugging the double mergesorts.
During the process, different functions were used to process lists that were
exactly the required size and lists which were not, which needed code to handle
edge-cases. However, the instruction count improvement was in the double digits,
and so these changes were removed in order to have lower instruction cache
misses, as well as to reduce the fill time of the branch predictors and reduce
the complexity and size of the algorithm.

@@

The instruction count of mergesort is $O(NlogN)$, and does not vary
according to input. Multi-mergesort should have a significantly higher
instruction count, once its threshold is hit, as should its variant, double
multi-mergesort.

@@

The number of instructions, shown in Figure \ref{merge instructions} is exactly
as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
\ref{merge level 2 misses} are also as expected, although the multi-mergesort
does not show the same improvement over tiled mergesort as in LaMarca's results.
Double multi-mergesort does show an improvement, but not a great improvement to
double tiled mergesort. It does, however, have a much flatter slope.

@@

@@

The difference between the actual and expected results of the tiled and
multi-mergesorts is another area to be investigated. While the results point to
similar conclusions about cache misses between the base, tiled and
multi-mergesorts, the relation these cache misses for direct mapped caches is
not expected. Investigating, and possibly fixing this would be interesting.

@@

A bug exists in the double tiled and multi-mergesorts, which causes an extra
merge step in some cases. More accurate results could be achieved by solving
this problem. Also, to take advantage of the level 1 cache, the double sorts were
sorted to 2048 keys first, then merged into larger sections. This number should
have been half that. See section \ref{bug4} for more details.

>>>>

\section{Double Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that $index(source) + index(auxiliary) =
C$.

@@

It is also possible to avoiding the copy back to memory. In the event of an odd
number of merge steps, the auxiliary array will contain the sorted array, which
needs to be copied back. This can be avoided by pre-sorting to make lists sized
two or eight. A large increase of instruction count was realized by sorting to
lists of length two, so a length of eight was used. When an even number was
required, four was used.

The final impriovement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since is the penalty
for level 1 misses is not great. For this reason, this sort is called Double
Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
in this report.

\section{Double Multi-mergesort}

Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the lists grow big, the overhead involved in this is
substantial. It is necessary to check each of the lists, and each key searched
would cause a cache miss, since each array segment is aligned, and map to
the same cache block.

@@

@@

The instruction count of mergesort is $O(NlogN)$, and does not vary
according to input. Multi-mergesort should have a significantly higher
instruction count once its threshold is hit.

@@

The number of instructions, shown in Figure \ref{merge instructions} is exactly
as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
Figure \ref{merge level 2 misses} are also as expected, although the
multi-mergesort does not show the same improvement over tiled mergesort as in
LaMarca's results.

@@

\section{Predictor results}

\afterpage{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort-0.eps}}
\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort-0.eps}}
\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-0.eps}}
\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
\label{Branch prediction performance for mergesorts}
\end{figure}
\newpage
}

%TODO This is a bit to rewrite

figure a shows Algorithm N, knuths natural merge. N3, N5 and N9 indicate parts of the flow chart. 'N3 i==j' is the second branch in N3.

The reuslts show that the seconsary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.

Algorithm S in graph B shows the same. I never really understood how it worked though.

Figure c shows base mergesort. This shows a lower number of branches and a higher precentage of correct predictions. The lower number of branches is probably due to the presort. there is a higher percentage of correct predictions, but the actual number of correct predictions in the same. This implies that doing the presort removes incorrect predictions (or, more to the point, moves incorrect predictions to another part of the algorithm). Note too, that both sides are used equally, which doesnt happen later. Two thirds of the branches are not-taken, which means loops execute twice before breaking. this is the same results that we found in quicksort later. Im sure knuth explains this somewhere in the fine print.

Figure d shows double tiled mergesort. The total number fo branches is a 10\% increase over the base mergesort. Not sure why this is. 15\% increase in correct predictions so this incrase into predictable range, which should be avoidable (according to what we guess in quicksort).

Figure e and f show the double mulitmergesort results. This sorts fully within the cache only, reducing the number of branches relating to this. The reason that the previous balance was so muhc in favour of forwards was that the final sort at the end was forwards. Here though, the arrays are left bitonically, so there is an even distribution betwwen forwards and backwards. There si still the bias towards left rather than right. Left comes first in the code, but this doesnt quite explain why this happens. However, the number fo misses each way is identical.

The total branches is a decrease of about 20\% on base mergesort. The number of mispredictions decreases by less than 10\%. Howver, all these branches are made up by the increase in branches due to the heap. The number of extra brnahc misses due to this is 26\% of the branches misses in base mergesort. In total, the num ber fo branches isa equivalent however. So no loss by doing this (except lots of extra instructions, as seen before).

Note that here and in the heaps, comparison 0 does nothing. THe compiler ought to trim it out, except that now it actually does something (increases the branch counter).

finally, in base mergesort, there is an inline presort used. For efficiency, an insertion sort is used later. However, there are 6.8 misses per key in the inline presort, which is less than the insertion sort does. Its not as useful for presorting to 8, however. THese results hold forwards and in reverse.

%TODO up to here

@@

@@

### 2005-05-26 paper/docs/ordernsquared.tex
make way for the new and improved fs

The branch prediction performance is not expected to be bad. Flow control
predictions are very straightforward, and should result in very few misses.
Comparison predictions, however, are very numerous. Each array traversal has an
average of $N/2$ comparisons, and approximately $log_2(N/2)$ of them should be
taken, while the rest are not-taken. The taken predictions will be misses,
while the rest should be hits, so it is expected that there will be $logN$
misses per key.

@@

@@

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse (see Section \ref{temporal locality}),
reducing its number of misses.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while \ref{Branch Simulation results for elementary sorts} shows the
branch misses per key.  Branch prediction performance is not surprising. Each
sort has a similar number of branches, with insertion sort having half as many
as selection and bubblesort, and shakersort being half-way between the two. The
number of misses for bubblesort and shakersort are huge; nearly one in four
comparisons are misses. As the data sets gets large, there are nearly 100,000
branch misses involved in sorting a single key.

@@

Shakersort doesn't have an opt-out-early clause like bubblesort, and as a result
would perform very badly for sorted sets. Adding this enhancement is simple, and
would give better results, which would also be more consistent with bubblesort.

>>>>

\label{selection branches logn}
The branch prediction performance is not expected to be bad. Flow control
predictions are very straightforward, and should result in very few misses.
Comparison predictions, however, are very numerous. Each array traversal has an
average of $N/2$ comparisons, and approximately $log_2(N/2)$ of them should be
taken, while the rest are not-taken. The taken predictions will be misses,
while the rest should be hits, so it is expected that there will be $logN$
misses per key.

@@

\section{Bubblesort2}
Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
implementation above ends early if it detects that the array is fully sorted.
It keeps a binary flag to indicate if there were any exchanges along the way.
Instead, bubblesort2 keeps track of the last exchange made, and starts the next
iteration from there. This means that any pockets of sorted keys are skipped
over, not just if they occur at the end. Bubblesort2 code is in Figure
\vref{Bubblesort2 code}

\code{Bubblesort2 code}{on2_bubblesort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the bubblesort2 should perform faster than the original in
all metrics. Instruction count, were there no conditions that allowed keys to be
skipped, would be the same, as no extra instructions are added. The flow of the
algorithm is the same, though there should be fewer iterations, especially at
the start, when iterations are longer, so cache accesses, and branch misses
should be reduced along with instruction count.

@@

\section{Shakersort2}
Shakersorts incorporates the same improvements over shakersort as bubblesort2
does over bubblesort, that is, it skips over pockets of sorted keys in both
directions, as well as ending early. The code for shakersort2 is in Figure
\vref{Shakersort2 code}.

\code{Shakersort2 code}{on2_shakersort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the improvements will make shakersort faster in all cases
than its original form.

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less. The improvement to
bubblesort in bubblesort2 have increased the instruction count, contrary to
predictions, and similarly in the shakersort2.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse (see Section \ref{temporal locality}),
reducing its number of misses. Shakersort2 is placed roughly between insertion
sort and shakersort, showing that ending early is helpful in its case. This does
not appear to be the case with bubblesort2 though, which performs identically to
bubblesort.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key.  Branch prediction performance is not surprising.
Each sort has a similar number of branches, with insertion sort having half as
many as selection and bubblesort, and shakersort being half-way between the
two. The number of misses for bubblesort and shakersort are huge; nearly one in
four comparisons are misses. As the data sets gets large, there are nearly
10,000 branch misses involved in sorting a single key. The improved bubblesort
had exactly the same number of misses and branches. The improved shakersort has
a lot less branches, but the same number of misses. This is because skipping
over sorted keys is predictable from the first branch miss.

@@

### 2005-05-26 paper/docs/radix.tex
make way for the new and improved fs

\section{Results}
\subsection{Test parameters}
Keys are to be divided into segments 8 bits long, resulting in four passes of
the array. The radix used is therefore $2^8 = 256$.

@@

The cache performance of radixsort is not expected to be good, however.
Quicksort and mergesort, for example, have very good temporal locality. Keys are
used repeatedly before being swapped out of the cache. In radixsort, however,
every key is used just once each time it is in the cache. When the array is larger than
the size of the cache, every key access will be a miss as a result, and every
new byte being sorted involves four passes over the array. In addition, if the
auxiliary array is pathologically aligned, then the copy back into the array
can result in thrashing.

@@

@@

The instruction count of radixsort, shown in Figure \ref{radix instructions} is
very low. It is lower than even quicksort for the smallest set, and it
flat-lines immediately (after amortisation of the loop overhead). The number of
cache misses in Figure \ref{radix level 2 misses} is exactly the same, until the
array no longer fits into the cache. The number of misses then leaps up to 2.5
per
key, which is roughly what was predicted.

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite 10 branches per key, there was less than .08 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected.

With the low number of branch misses, a much better overall performance is
expected than is achieved. When in the cache, it performs only slightly worse
than quicksort, but once the data set grows larger, the gap widens.  However,
the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
never increase beyond 350 cycles per key, meaning that for larger sets it will
perform significantly better than quicksort. The reason for the high number of
cycles is that radixsort was not optimised. Possible optimisations are discussed
below, and should significantly reduce the number of cache misses, the main
cause of radixsorts poor performance.

@@

\section{Future Work}
\label{radix future work}
Many improvements exist to radixsort, and some are suggested by Sedgewick and
LaMarca. Some of these improvements help to lower cache misses and instruction
count, both of which slow down radixsort significantly. LaMarca suggested
that while the current byte is being put in place, the counting of the next byte
can be done. This increases the memory requirements, but this is not
significant.

The first improvement is to reduce the number of runs across the array per step.
At the moment, there is one step to count, one to copy, and one to write back
(causing potentially two misses per write). LaMarca suggests counting for the next
step in the previous count. In fact, all the counting can be done in a single
pass at the start, instead of four passes at various stages, decreasing instruction
count and cache misses. The downside is the fourfold increase in memory required
for the arrays of counts, but this is small compared to the main and auxiliary
arrays.

The step to copy the array back can also be removed. In mergesort, copying back
to the source array was changed to copying back and forth between the source and 
auxiliary arrays. The same could be potentially done with radixsort, without any
increase in complexity. The number off steps in likely to be even, and is known
at compile time, so the number of times the copy step is done could be reduced
to zero. If the number of steps is odd, which is exceptionally unlikely when
sorting integers, then one copy at the end would be required.

Combining other sorts with radixsort could also be possible. An optimisation to
mergesort was to do a k-way merge at the end, but this dramatically increased
the instruction count. Using a type of radixsort here would be complicated, but
may be possible. Alternatively, sorting inside the cache using radixsort is far
faster than merging. The k-way merge could then be performed at the end. In
practice, though, radixsort is far faster than mergesort, has a lower
instruction count, and less cache misses. While these steps could improve
mergesort, they are more likely to slow radixsort, with no advantage.

@@

Additional performance improvements could possibly be gained using a different
radix. This radix was chosen simply because Sedgewick used it. Another radix may
lead to better results.

>>>>

\subsection{Cache Radixsort}
Many improvements exist to radixsort, and some are suggested by Sedgewick and
LaMarca. Some of these improvements help to lower cache misses and instruction
count, both of which slow down radixsort significantly. LaMarca suggested
that while the current byte is being put in place, the counting of the next byte
can be done. This increases the memory requirements, but this is not
significant.

The first improvement is to reduce the number of runs across the array per
step.  In the \n{base radixsort}, presented above, there is one step to count,
one to copy, and one to write back (which accesses both arrays). LaMarca
suggests counting for the next step in the previous count. In fact, all the
counting can be done in a single pass at the start, instead of four passes at
various stages, decreasing instruction count and cache misses. The downside is
the fourfold increase in memory required for the arrays of counts, with no
benefits.

The step to copy the array back can also be removed. In mergesort, copying back
to the source array was changed to copying back and forth between the source and 
auxiliary arrays. The same could be potentially done with radixsort, without any
increase in complexity. The number of steps is likely to be even, and is known
at compile time, so the number of times the copy step is done could be reduced
to zero. If the number of steps is odd, which is unlikely when sorting
integers, then one copy at the end would be required. This is not needed in the
implementation presented here.

Cache radixsort should traverse the array only nine times: Once at the start to
count the least significant bits; six times as it copies using the three least
significant bytes (counting the next least significant byte at the same time),
and a final two as it copies using the most significant byte.

\section{Results}
\subsection{Test parameters}
Keys are to be divided into segments 8 bits long, resulting in four sorting
steps to completely sort the array. The radix used is therefore $2^8 = 256$.

@@

The cache performance of base radixsort is not expected to be good, however.
Quicksort and mergesort, for example, have very good temporal locality. Keys are
used repeatedly before being swapped out of the cache. In radixsort, however,
every key is used just once each time it is in the cache. When the array is larger than
the size of the cache, every key access will be a miss as a result, and every
new byte being sorted involves four passes over the array. In addition, if the
auxiliary array is pathologically aligned, then the copy back into the array
can result in thrashing.

@@

Due to the reduced copying, cache radixsort should have a lower instruction
count and fewer cache misses that base radixsort. There should be nine passes
over the array, leading to just over one miss per key, as opposed to two misses
per key with base radixsort. The complexity and branch prediction results
should remain the same.

@@

The instruction count of base radixsort, shown in Figure \ref{radix instructions} is
very low. It is lower than even quicksort for the smallest set, and it
flat-lines immediately (after amortisation of the loop overhead). The number of
cache misses in Figure \ref{radix level 2 misses} is exactly the same, until
the array no longer fits into the cache. The number of misses then leaps up to
2.5 per key, which is roughly what was predicted.

Cache radixsort performs as predicted aswell. By sorting keys between two
arrays instead of copying a partially sorted array back, the instruction count
has been reduced by more than 10\%, and the number of level 2 cache misses
have been more than halved, to just over one miss per key.

The number of branch misses, shown in Figure \ref{radix branch misses} was also
extremely low. Despite 10 branches per key, there was less than .08 of a branch
miss per key. The number of branches itself is quite low, as several of the
sorts considered had nearly that number of misses. However, it is not
unexpected. The number of branches reduces when copying back is removed, and
while this reduces the number of mispredictions, there are so few to begin with
that this result is not significant.

With the low number of branch misses, a much better overall performance is
expected than is achieved. When in the cache, it performs only slightly worse
than quicksort, but once the data set grows larger, the gap widens.  However,
the results, shown in Figure \ref{radix cycles}, indicate that radixsort will
never increase beyond 350 cycles per key, meaning that for larger sets it will
perform significantly better than quicksort. Cache radixsort never increases
beyond 300 cycles, though it has relatively more cycles when the arrays are
smaller.

The most curious result is that all simulations show that cache radixsort
performing better than the base radixsort, yet the Pentium 4 results show that
base radixsort performs better for smaller arrays.

@@

\section{Future Work}
\label{radix future work}
Combining other sorts with radixsort could also be possible. An optimisation to
mergesort was to do a k-way merge at the end, but this dramatically increased
the instruction count. Using a type of radixsort here would be complicated, but
may be possible. Alternatively, sorting inside the cache using radixsort is far
faster than merging. The k-way merge could then be performed at the end. In
practice, though, radixsort is far faster than mergesort, has a lower
instruction count, and less cache misses. While these steps could improve
mergesort, they are more likely to slow radixsort, with no advantage.

@@
